---
title: "Rapid access to Mass-Spec data (RaMS)"
author: ""
date: ""
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RaMS)
par(mar=c(4.1, 4.1, 0.1, 0.1))
```

## Overview

`RaMS` is a lightweight package that provides rapid and tidy access to 
mass-spectrometry data. This package is *lightweight* because it's built from
the ground up rather than relying on an extensive network of external libraries.
No Rcpp, no Bioconductor, no long load times and strange startup warnings. 
Just XML parsing provided by `xml2` and data handling
provided by `data.table`. Access is *rapid* because an absolute minimum of data
processing occurs. Unlike other packages, `RaMS` makes no assumptions about
what you'd like to do with the data and is simply providing access to the
encoded information in an intuitive and R-friendly way. Finally, the access is
*tidy* in the philosophy of [tidy data](https://r4ds.had.co.nz/tidy-data.html).
Tidy data neatly resolves the ragged arrays that mass spectrometers produce and
plays nicely with other [tidy data packages](https://www.tidyverse.org/).

## Installation

Until `RaMS` is on CRAN, the easiest way to install is via `devtools`:

```{r, echo=TRUE, eval=FALSE}
devtools::install_github("wkumler/RaMS")

library(RaMS)
```

## Usage

There's only one main function in `RaMS`: the aptly named `grabMSdata`. This
function accepts the names of mass-spectrometry files as well as the data you'd
like to extract (e.g. MS1, MS2, BPC, etc.) and produces a list of data tables.
Each table is intuitively named within the list and formatted tidily:

```{r, results='hide'}
demo_dir <- system.file("extdata", package = "RaMS")
msdata_files <- list.files(demo_dir, pattern = "mzML", full.names = TRUE)
output <- grabMSdata(files = msdata_files, grab_what = c("TIC", "MS1"))
```

```{r}
knitr::kable(head(output$MS1))

knitr::kable(head(output$TIC))
```

This means that the basic R functions work exactly as we expect them to, no
new method definition or memorization necessary:

```{r}
str(output)

summary(output$MS1)

# Outputs are data.tables so we can use their intuitive indexing on column name
first_file_data <- output$TIC[filename==basename(msdata_files[1])]
plot(first_file_data$rt, first_file_data$int, type = "l")
```

And of course, the tidy data format means that it plays nicely with every other
tidy data package.

```{r}
# Plot some chromatograms with ggplot2
library(ggplot2)
ggplot(output$TIC) + geom_line(aes(x = rt, y=int, color=filename)) + theme(legend.position="top")
ggplot(output$TIC) + geom_line(aes(x = rt, y=int)) +
  facet_wrap(~filename, scales = "free_y", ncol = 1) +
  labs(x="Retention time (min)", y="Intensity")

# Manually calculate a BPC with dplyr
suppressMessages(library(dplyr))
manual_TIC <- output$MS1 %>%
  group_by(rt, filename) %>%
  summarise(manual_int=sum(int), .groups="drop") %>%
  left_join(output$TIC, by = c("rt", "filename"))

# Return to base to plot TIC comparison
par(pty="s")
plot(manual_TIC$int, manual_TIC$manual_int, 
     ylab = "Manually calculated TIC",
     xlab = "TIC read from file")
grid()
corr <- cor(manual_TIC$int, manual_TIC$manual_int)
legend("topleft", legend=paste("Correlation =", corr))
```

## File types

RaMS is currently limited to the modern **mzML** data format and the slightly 
older **mzXML** format. Tools to convert data from other formats are available
through [Proteowizard](http://proteowizard.sourceforge.net/tools.shtml)'s 
`msconvert` tool. Data can, however, be gzip compressed (file ending .gz) and
this compression actually speeds up data retrieval significantly as well as
reducing file sizes.

Currently, `RaMS` also handles only MS<sup>1</sup> and MS<sup>2</sup> data. This should be easy enough
to expand in the future, but right now I haven't observed a demonstrated need
for higher fragmentation level data collection.
