---
title: "Intro to RaMS"
output: html_document
vignette: >
  %\VignetteIndexEntry{RaMS-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "80%",
  fig.align = 'center',
  fig.height = 3
)
```

Welcome to RaMS! This vignette is designed to provide a slew of examples using 
the package at various levels of complexity. Let's jump right in.

If you have your own data, feel free to load it here. If not, there's a couple
example small files you're welcome to use in the "inst" folder. I'll be using
these throughout, and more information is available in the addendum below.

```{r findfiles, message=FALSE}
library(RaMS)
library(data.table)
library(tidyverse)
library(plotly)

# Locate the file directory
msdata_dir <- system.file("extdata", package = "RaMS")

# Identify the files of interest
data_files <- list.files(msdata_dir, pattern = "Full.*mzML", full.names = TRUE)

# Check that the files identified are the ones expected
basename(data_files)
```

There's only one function to worry about in RaMS: the aptly named `grabMSdata`.
This function has a couple arguments with sensible defaults, but you'll always
need to tell it two things: one, which files you'd like to process; and two,
the data you'd like to obtain from those files.

Let's start simple, with a single file and the most basic information about it.

## Basic RaMS usage

### TICs, BPCs, and metadata

A TIC reports the total intensity measured by the mass analyzer during each 
scan, so the data is parsed into two columns: retention time (rt) and intensity
(int). This makes it easy to read and simple to plot:

```{r loadTIC}
single_file <- data_files[1]

file_data <- grabMSdata(single_file, grab_what = "TIC")

knitr::kable(head(file_data$TIC, 3))
```

Since we asked for a single thing, the TIC, our `file_data` object is a list
with a single entry: the TIC. Let's plot that data:

```{r headerTIC}
par(mar=c(4.1, 4.1, 0.1, 0.1))
plot(file_data$TIC$rt, file_data$TIC$int, type = "l")
```

Simple enough!

A BPC is just like a TIC except that it records the *maximum* intensity measured,
rather than the sum of all intensities. This data is also collected by the mass
analyzer and doesn't need to be calculated.

```{r loadBPC}
file_data <- grabMSdata(single_file, grab_what = "BPC")
```

Since the data is parsed in a ["tidy" format](https://r4ds.had.co.nz/tidy-data.html), 
it plays nicely with popular packages such as `ggplot2`. Let's use that to plot
our BPC instead of the base R plotting system:

```{r plotBPC}
ggplot(file_data$BPC) + geom_line(aes(x=rt, y=int))
```

The advantages of tidy data and `ggplot` become clear when we load more than
one file at a time because we can group and color by the third column, the
name of the file from which the data was read. Here I've also enabled a progress
bar with the argument `verbosity="minimal"` because it's nice to see progress
when multiple files are being read.

```{r loadmultiBPC}
file_data <- grabMSdata(data_files, grab_what = "BPC", verbosity = "minimal")

ggplot(file_data$BPC) + geom_line(aes(x=rt, y=int, color=filename))
```

And of course, this means that all of `ggplot`'s aesthetic power can be brought
to your chromatograms as well, so customize away!

```{r ggplotRaMS, dev.args=list(png  = list(type = "cairo"))}
ggplot(file_data$BPC) + 
  geom_line(aes(x=rt, y=int, color=filename), lwd=1.2) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text = element_text(color = "#AA0000"),
        axis.title = element_text(family = "serif"),
        plot.title = element_text(face = "bold")) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x="Retention time (min)", y="Intensity", 
       title = "My chromatogram", color="File names:")
```

`RaMS` also provides some basic file metadata extraction capability, although
the focus for this package is on the actual data and other MS packages handle
file metadata much more elegantly. This is one area where there are major
differences between mzML and mzXML file types - the mzXML file type simply
doesn't encode as much metadata as the mzML filetype, so `RaMS` can't extract
it.

```{r}
mzML_file <- list.files(msdata_dir, pattern = "Full.*mzML", full.names = TRUE)[1]
grabMSdata(mzML_file, grab_what = "metadata")

mzXML_file <- list.files(msdata_dir, pattern = "Full.*mzXML", full.names = TRUE)[1]
grabMSdata(mzXML_file, grab_what = "metadata")
```

### Adding a column: MS1 data

MS1 data can be extracted just as easily, by supplying "MS1" to the `grab_what`
argument of `grabMSdata` function.

```{r}
file_data <- grabMSdata(data_files, grab_what = "MS1")

knitr::kable(head(file_data$MS1, 3))
```

So we've now got the *mz* column, corresponding to the mass-to-charge ratio 
(*m/z*) of an ion. This means that we can now filter our data for specific
masses and separate out molecules with different masses.

Note that this also makes the data much larger in R's memory - so don't go
loading hundreds of files simultaneously. If that's necessary, check out the
section below on saving space.

Because `RaMS` returns [data.tables](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) rather than normal `data.frame`s, indexing
is super-fast and a bit more intuitive than with base R. Below, I also use the
`pmppm` function from `RaMS` to produce a mass range from an initial mass and
spectrometer accuracy (here, 5 parts-per-million).

```{r}
adenine_mz <- 136.06232

adenine_data <- file_data$MS1[mz%between%pmppm(adenine_mz, ppm=5)]

ggplot(adenine_data) + geom_line(aes(x=rt, y=int, color=filename))
```

This makes it easy to grab the data for multiple compounds of interest with a
simple loop, provided here by the `purrr` package of the tidyverse:

```{r}
masses_of_interest <- c(adenine=136.06232, valine=118.0865, homarine=138.055503)

mass_data <- imap(masses_of_interest, function(mz_i, name){
  cbind(file_data$MS1[mz%between%pmppm(mz_i, ppm=5)], name)
}) %>% rbindlist()

ggplot(mass_data) + 
  geom_line(aes(x=rt, y=int, color=filename)) + 
  facet_wrap(~name, ncol = 1, scales = "free_y")
```

### Moving along: MS2 data

`RaMS` also handles MS<sup>2</sup> data elegantly. Request it with the "MS2" 
option for `grab_what`, although it's often a good idea to grab the MS1 data
alongside.

```{r}
DDA_file <- list.files(msdata_dir, pattern = "DDA.*mzML", full.names = TRUE)
DDA_data <- grabMSdata(DDA_file, grab_what = c("MS2"))
knitr::kable(head(DDA_data$MS2, 3))
```

DDA data can be plotted nicely with `ggplot2` as well. Typically it makes sense
to filter for a precursor mass, then render the fragments obtained.

```{r}
betaine_mass <- 118.0865

betaine_MS2 <- DDA_data$MS2[premz%between%pmppm(betaine_mass, 5)]
betaine_MS2$int <- betaine_MS2$int/max(betaine_MS2$int)*100

ggplot(betaine_MS2) +
  geom_point(aes(x=fragmz, y=int)) +
  geom_segment(aes(x=fragmz, xend=fragmz, y=int, yend=0)) +
  labs(x="Fragment m/z", y="Relative intensity (%)")
```


This is also the perfect place to enable some interactivity with packages such
as `plotly`, making data exploration not only simple but also enjoyable. 

```{r plotly, warning=FALSE}
data_files <- list.files(msdata_dir, pattern = "mzML", full.names = TRUE)
file_data <- grabMSdata(data_files, grab_what = c("MS1", "MS2"))

clean_MS2 <- file_data$MS2 %>% 
  filter(premz%between%pmppm(betaine_mass)) %>%
  group_by(rt) %>%
  arrange(desc(int)) %>%
  summarise(frags=paste(
    paste(round(fragmz, digits = 3), round(int), sep = ": "), collapse = "\n"),
    .groups="drop"
  )
file_data$MS1 %>% 
  filter(mz%between%pmppm(betaine_mass)) %>%
  filter(!str_detect(filename, "DDA")) %>%
  plot_ly() %>%
  add_trace(type="scatter", mode="lines", x=~rt, y=~int, color=~filename,
            hoverinfo="none") %>%
  add_trace(type="scatter", mode="markers", x=~rt, y=0,
            text=~frags, hoverinfo="text", showlegend=FALSE,
            marker=list(color="black"), data = clean_MS2) %>%
  layout(annotations=list(x=min(clean_MS2$rt), y=0, 
                          text="Mouse over to see\nMSMS fragments"))
```



Easy access to MS<sup>2</sup> data also allows us to rapidly perform 
simple operations such as searching for a
specific fragment mass. For example, if we know that glycine betaine typically
produces a fragment with a mass of 58.0660, we simply subset the MS<sup>2</sup> data
for fragments in a range around that mass:

```{r}
betaine_frag_mz <- 58.0660
knitr::kable(head(DDA_data$MS2[fragmz%between%pmppm(betaine_frag_mz, ppm = 5)]))
```

We find that there's not only glycine betaine that produces that fragment,
but also another compounds with a mass of 212.0951. Fragments like this can then
be searched manually in online databases or, since the data is already in R,
passed to a script that automatically searches them.

Similarly, we can easily search instead for neutral losses with this method. If
we suspect other molecules are producing a similar neutral loss as glycine
betaine:

```{r}
betaine_mass <- 118.0865
betaine_neutral_loss <- betaine_mass - betaine_frag_mz

DDA_data$MS2 <- mutate(DDA_data$MS2, neutral_loss=premz-fragmz)
DDA_data$MS2[neutral_loss%between%pmppm(betaine_neutral_loss, ppm = 5)] %>%
  head() %>% knitr::kable()
```

We can again confirm our suspicions that there's another molecule with
a similar neutral loss: one with a mass of 138.0549.

## Advanced RaMS usage

### Saving space: EICs

### Fine control: grabMzmlData, grabMzxmlData

### Speeding things up

### The nitty-gritty details

## Addendum: About the data
